---
title: "P8106_HW5_sp4436"
author: "Sukhman Parhar"
date: "2026-05-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(caret)
library(tidymodels)
library(kernlab)
library(e1071)
library(ISLR)
library(factoextra)
library(gridExtra)
```

### Question 1

#### Clean the data

```{r}
# Load in the data 
auto <- read_csv("auto.csv") %>% 
  na.omit() %>% 
  as.data.frame()

# Remove cars with 3 cylinders
auto <- auto[auto$cylinders != 3, ]

# Convert Outcome (mpg_cat) to factor
auto$mpg_cat <- factor(auto$mpg_cat)

# Convert Origin to factor 
auto$origin <- as.factor(auto$origin)

#summary(auto)
```

#### Split the data

```{r}
set.seed(2)

split <- initial_split(auto, prop = 0.7)

train_data <- training(split)
test_data  <- testing(split)

# set up cross validation control
ctrl <- trainControl(method = "cv", number = 10,
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE)
```

#### Part A: Fit a support vector classifier (SVC) to the training data. What are the training and
test error rates?

```{r}
set.seed(2)

# Support vector classifier: linear kernel
linear.tune <- tune.svm(mpg_cat ~ .,
                        data = train_data,
                        kernel = "linear",
                        cost = exp(seq(-5, 2, len = 50)),
                        scale = TRUE)

plot(linear.tune, highlight = TRUE, xTrans = log)
best.linear <- linear.tune$best.model

pred.linear.train <- predict(best.linear, newdata = train_data)
pred.linear.test  <- predict(best.linear, newdata = test_data)

mean(pred.linear.train != train_data$mpg_cat)

svc.test.error <- mean(pred.linear.test != test_data$mpg_cat)
svc.test.error
```

##### Training and test error
Training error = 0.06642066
Test error = 0.08547009

#### Part B: Fit a support vector machine (SVM) with a radial kernel to the training data. What
are the training and test error rates?

##### Tuning Grid

```{r}
svmr.grid = expand.grid(C = exp(seq(1, 7, len = 50)),
                        sigma = exp(seq(-10, -2, len = 20)))
```

##### Tune

```{r}
set.seed(2)

svmr.fit = train(mpg_cat ~ . ,
                 data = train_data,
                 method = "svmRadialSigma",
                 tuneGrid = svmr.grid,
                 trControl = ctrl)

myCol = rainbow(25)
myPar = list(superpose.symbol = list(myCol),
             superpose.line = list(col = myCol))

plot(svmr.fit, highlight = TRUE, par.settings = myPar)

svmr.fit$bestTune

confusionMatrix(svmr.fit)

pred.train.rad <- predict(svmr.fit, newdata = train_data)
pred.test.rad  <- predict(svmr.fit, newdata = test_data)

train.error.rad <- mean(pred.train.rad != train_data$mpg_cat)
test.error.rad  <- mean(pred.test.rad  != test_data$mpg_cat)

train.error.rad
test.error.rad
```

##### Training and test error
Training error = 0.05535055
Test error = 0.05982906

#### Part C: Compare the performance of the above models with boosting. Do SVC or SVM
provide better predictive performance on this dataset?

```{r}
gbmA.grid = expand.grid(n.trees = c(100, 200, 500, 1000, 2000, 5000, 10000),
                        interaction.depth = 1:5,
                        shrinkage = c(0.005, 0.007, 0.009),
                        n.minobsinnode = 10)

set.seed(2)

gbmA.fit = train(mpg_cat ~ . ,
                 train_data,
                 tuneGrid = gbmA.grid,
                 trControl = ctrl,
                 method = "gbm",
                 distribution = "adaboost",
                 metric = "Accuracy",
                 verbose = FALSE)

ggplot(gbmA.fit, highlight = TRUE)

pred.gbm.train <- predict(gbmA.fit, newdata = train_data)
pred.gbm.test  <- predict(gbmA.fit, newdata = test_data)

train.error.gbm <- mean(pred.gbm.train != train_data$mpg_cat)
test.error.gbm  <- mean(pred.gbm.test  != test_data$mpg_cat)

train.error.gbm
test.error.gbm
```

```{r}
svc.test.error
test.error.rad
test.error.gbm
```

